Dark Control:\\ Towards a Unified Account of Default Mode Function\\ by Markov Decision Processes

### 2017-12-08 paper/paper.tex
reread part 4

What has early been described as the ``stream of consciousness''
in psychology \citep{james1890principles}
found a potential neurobiological manifestation
in the DMN
\citep{shul1997, raichle2001pnas}.
We propose that this set of some of the most advanced regions in
the association cortex \citep{mesulam1998sensation, margulies2016situating}
are responsible for higher-order control of human behavior.
Our functional account follows the notion of
``a hierarchy of brain systems with
the DMN at the top and the salience and dorsal attention systems
at intermediate levels, above thalamic and primary sensory
cortex'' \citep{carhart2010default}.

@@

Psychologically,
the more the ongoing executed task is unknown and unpracticed,
the less stimulus-independent thoughts occur
\citep{filler1973daydreaming, teasdale1995stimulus, christoff2016mind}.
Conversely,
it has been empirically shown that, the more the world is easy to foresee,
the more human mental activity
becomes detached from the actual sensory environment
\citep{antrobus1966studies, pope1978regulation, mason2007, weiss2006}.
Without requiring explicit awareness,
these ``offline'' processes may contribute
to optimizing control of the organism.
We formalize
a \textit{policy matrix} to capture the space of possible actions that
the agent can perform
on the environment given the current state. A \textit{value function}
maps environmental objects and events (i.e., states) to expected rewards.
Switching between states reduces to a sequential processing model.
Informed by outcomes of performed actions, neural computation reflected in
DMN dynamics could be constantly shaped by prediction error through feedback loops.
This computational account of DMN will be described in the mathematical framework of Markov Decision Processes (MDPs). MDPs indeed address specifically the case of decision making in stochastic context with reward feedback.

@@

In artificial intelligence and machine learning, a popular computational model for
multi-step decision processes in such an environment are
\textit{Markov decision processes} (MDPs) \citep{sutton1998reinforcement}.
% For the remainder of this paper, for the sake of simplicity
% we use the abbreviation ``MDP'' to refer to the special case
% of \textit{Partially Observable Markov Decision Processes}.
An MDP operationalizes a sequential decision process
in which it is assumed that environment dynamics are determined by a Markov process,
but the agent cannot directly observe the underlying state.
Instead, the agent tries to optimize a \textit{subjective} reward
signal (i.e., likely to be different for another agent in the same state),
by maintaining probability distributions over actions according to their expected utility.
% This is performed alongside each transition in the state space of the environment
%  and maintains a probability distribution over the set of possible
% states.
% % , based on a set of observations and observation probabilities.
This is a minimal set of assumptions that can be made about an environment faced by an agent engaged in interactive learning.

@@

\subsubsection{Accumulated rewards and policies}
The behavior of the agent is governed by a \textit{policy}, which maps states of the world
to probability distributions over candidate actions.
Starting at time $t=0$,
following a policy $\pi$ generates a trajectory of action choices:
\begin{eqnarray*}
  \begin{split}
    \textbf{choose action: }a_0 &\sim \pi(a|s_0)\\
    \textbf{ observe transition: }s_1 &\sim p(s|s_0,a_0)\textbf{ and collect reward }R_0 = r(s_0, a_0, s_1)\\
    \textbf{choose action: }a_1 &\sim \pi(a|s_1)\\
    \textbf{ observe transition: }s_2 &\sim p(s|s_1,a_1), \textbf{ and collect reward }R_1 = r(s_1, a_1,s_2)\\
    \vdots\\
    \textbf{choose action: }a_{t} &\sim \pi(a|s_{t})\\
    \textbf{ observe transition: }s_{t+1} &\sim p(s|s_{t},a_{t}), \textbf{ and collect reward }R_{t} = r(s_{t}, a_{t},s_{t+1})\\
    \vdots
  \end{split}
\end{eqnarray*}
We assume time-invariance
in that we expect the dynamics of the process
to be equivalent over sufficiently long time windows of equal length (i.e., stationarity).
Since an action executed in the present moment might have repercussions in the far future, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s, a)$, but a
\textit{cumulative reward} estimate which takes into account expected reward
from action choices in the future.
A common approach to
modeling this accumulation is the time-discounted cumulative reward% function
\begin{equation}
  \label{eq:cumr}
  G^\pi = \sum_{t=0}^{\infty}\gamma^{t}R_t = R_0 + \gamma R_1 + \gamma^2 R_2 + \ldots + \gamma^tR_t + \ldots
\end{equation}
This random variable\footnote{Random as it depends both on the environment's dynamic and the
  policy $\pi$ being played (which can be stochastic). The exponential delay discounting function used here refere to the usual formulation in the field of reinforcement learning but psychological experiments may also reveal other discounting regimes \citep{green2004discounting}}  measures the cumulative reward of
following an action policy $\pi$.
%
Note that value buffering may be realized in the vmPFC.
This DMN region has direct connections to
to the NAc, known to be involved in reward evaluation.

>>>>

What has early been described as the ``stream of consciousness''
in psychology \citep{james1890principles}
found a potential neurobiological manifestation
in the DMN
\citep{shul1997, raichle2001pnas}.
We propose that this set of some of the most advanced regions in
the association cortex \citep{mesulam1998sensation, margulies2016situating}
are responsible for higher-order control of human behavior.
Our functional account follows the notion of
``a hierarchy of brain systems with
the DMN at the top and the salience and dorsal attention systems
at intermediate levels, above thalamic and unimodal sensory
cortex'' \citep{carhart2010default}.

@@

Psychologically,
the more the ongoing executed task is unknown and unpracticed,
the less stimulus-independent thoughts occur
\citep{filler1973daydreaming, teasdale1995stimulus, christoff2016mind}.
Conversely,
it has been empirically shown that, the more the world is easy to foresee,
the more human mental activity
becomes detached from the actual sensory environment
\citep{antrobus1966studies, pope1978regulation, mason2007, weiss2006}.
Without requiring explicit awareness,
these ``offline'' processes may contribute
to optimizing control of the organism.
We formalize
a \textit{policy matrix} to capture the space of possible actions that
the agent can perform
on the environment given the current state. A \textit{value function}
maps environmental objects and events (i.e., states) to expected rewards.
Switching between states reduces to a sequential processing model.
Informed by outcomes of performed actions, neural computation reflected in
DMN dynamics could be constantly shaped by prediction error through feedback loops.
The present computational account of DMN function will be described in the mathematical framework of Markov Decision Processes (MDP). MDPs specifically formalize decision making in stochastic contexts with reward feedback.

@@

In artificial intelligence and machine learning, a popular computational model for
multi-step decision processes in such an environment are
MDPs \citep{sutton1998reinforcement}.
% For the remainder of this paper, for the sake of simplicity
% we use the abbreviation ``MDP'' to refer to the special case
% of \textit{Partially Observable Markov Decision Processes}.
An MDP operationalizes a sequential decision process
in which it is assumed that environment dynamics are determined by a Markov process,
but the agent cannot directly observe the underlying state.
Instead, the agent tries to optimize a \textit{subjective} reward
signal (i.e., likely to be different for another agent in the same state),
by maintaining probability distributions over actions according to their expected utility.
% This is performed alongside each transition in the state space of the environment
%  and maintains a probability distribution over the set of possible
% states.
% % , based on a set of observations and observation probabilities.
This is a minimal set of assumptions that can be made about an environment faced by an agent engaged in interactive learning.

@@

\subsubsection{Accumulated rewards and policies}
The behavior of the agent is governed by a \textit{policy}, which maps states of the world
to probability distributions over candidate actions.
Starting at time $t=0$,
following a policy $\pi$ generates a trajectory of action choices:
\begin{eqnarray*}
  \begin{split}
    \textbf{choose action: }a_0 &\sim \pi(a|s_0)\\
    \textbf{ observe transition: }s_1 &\sim p(s|s_0,a_0)\textbf{ and collect reward }R_0 = r(s_0, a_0, s_1)\\
    \textbf{choose action: }a_1 &\sim \pi(a|s_1)\\
    \textbf{ observe transition: }s_2 &\sim p(s|s_1,a_1), \textbf{ and collect reward }R_1 = r(s_1, a_1,s_2)\\
    \vdots\\
    \textbf{choose action: }a_{t} &\sim \pi(a|s_{t})\\
    \textbf{ observe transition: }s_{t+1} &\sim p(s|s_{t},a_{t}), \textbf{ and collect reward }R_{t} = r(s_{t}, a_{t},s_{t+1})\\
    \vdots
  \end{split}
\end{eqnarray*}
We assume time-invariance
in that we expect the dynamics of the process
to be equivalent over sufficiently long time windows of equal length (i.e., stationarity).
Since an action executed in the present moment might have repercussions in the far future, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s, a)$, but a
\textit{cumulative reward} estimate which takes into account expected reward
from action choices in the future.
A common approach to
modeling this accumulation is the time-discounted cumulative reward% function
\begin{equation}
  \label{eq:cumr}
  G^\pi = \sum_{t=0}^{\infty}\gamma^{t}R_t = R_0 + \gamma R_1 + \gamma^2 R_2 + \ldots + \gamma^tR_t + \ldots
\end{equation}
This random variable

  \footnote{Random as it depends both on the environment's dynamics and the
  policy $\pi$ being executed. The exponential delay discounting function used here refers to the usual formulation in the field of reinforcement learning, although psychological experiments may also reveal other discounting regimes \citep{green2004discounting}}

measures the cumulative reward of
following an action policy $\pi$.
%
Note that value buffering may be realized in the vmPFC.
This DMN region has direct connections to
to the NAc, known to be involved in reward evaluation.
